Finally Tagless, Partially Evaluated

### 2008-11-05 tagless/tagless-final.tex
Finish addressing Olivier's comments

Most interpreters suffer from various kinds of overhead, making it less
efficient to run object programs via the metalanguage than to implement
the object language directly on the machine running the metalanguage
\citep{jones-partial}.  Two major sources of overhead are dispatching on
the syntax of object terms and checking the types of object values.  If
the metalanguage supports code generation
\citep{nielson-two-level,nielson-automatic,gomard-partial,bawden-quasiquotation,taha-sound},
then the embedding can avoid the dispatching overhead by compiling
object programs, that is, by specializing an interpreter to object
programs \citep{futamura-partial}.  Specializing an interpreter is thus
a promising way to build a DSL\@.  However, the type-checking overhead
remains, especially if the object language and the metalanguage both
have a sound type system.  This overhead has motivated much work on
typed specialization
\citep{danvy-simple,Danvy-tagging-encoding,hughes-type,Birkedal-PE-ML,taha-tag,Makholm-TagElim}
and type systems (see~\S\ref{s:fancier}).

@@

This paper shows how to eliminate type-checking overhead, whether in the
context of code generation and whether in the presence of dispatching
overhead.  We use metalanguage types, without such fancy features as
generalized algebraic data types (GADTs) or dependent types, to rule out
ill-typed object terms statically,
thus speeding up interpretation and assuring that our interpreters
do not get stuck.  We illustrate the problem of type-checking overhead
in this section using a simple evaluator as example.  We apply our
solution first to evaluation, then to code\hyp generation tasks such as
partial evaluation.

@@

Conversely, the type system of the metalanguage checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checkers also flag syntactically invalid
object terms, such as if we forget |app| somewhere above.
\ccshan{Make the point that the error messages are nice because we also
metacircularly take advantage of any niceness in the error messages of
the metalanguage implementation!}

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, we unfold |fix| rather than residualize it.
A conservative alternative is to unfold recursion only once, then residualize:
\begin{code}
let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
\end{code}
Many sophisticated approaches have been developed to decide how much to unfold
\citep{Jones-Mix,jones-partial}, but this issue is 
orthogonal to our presentation.
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{BondorfDanvy,SwadiTahaKiselyovPasalic2006}.
\ccshan{Fix the citations in the preceding sentence.}

@@

Our partial evaluator owes much to
\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducers for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen,Mogensen-SelfApplicable} also
build static and dynamic terms in tandem.  However, they
build a static term for every object term, even a bound variable,
so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remain untyped.  
In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.
The idea of generating static and dynamic components alongside each
other is found in an early form in Schism \citep[\S5]{consel-schism-93}. 
\ccshan{Refer to PE tradition and (earlier, even) the category-theory
concept of \emph{gluing}.}

@@

\ccshan{The following should be far from the first mention of optimality
in this paper.}
\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
apply HOAS to typed self\hyp interpretation.  However, their source
language is untyped.  Therefore, their object\hyp term encoding has
tags, and their interpreter can raise run-time errors.
Nevertheless, HOAS lets the partial
evaluator remove all the tags. In contrast, our object encoding and
interpreters do not have tags to start with and obviously cannot
raise run-time errors.

>>>>

Most interpreters suffer from various kinds of overhead, making it less
efficient to run object programs via the metalanguage than to implement
the object language directly on the machine running the metalanguage
\citep{jones-partial}.  Two major sources of overhead are dispatching on
the syntax of object terms and tagging the types of object values.  If
the metalanguage supports code generation
\citep{nielson-two-level,nielson-automatic,gomard-partial,bawden-quasiquotation,taha-sound},
then the embedding can avoid the dispatching overhead by compiling
object programs, that is, by specializing an interpreter to object
programs \citep{futamura-partial}.  Specializing an interpreter is thus
a promising way to build a DSL\@.  However, the tagging overhead
remains, especially if the object language and the metalanguage both
have a sound type system.
The quest to remove all interpretive overhead, in particular by
specializing the interpreter using a \emph{Jones-optimal} partial
evaluator \citep{jones-partial}, has motivated much work on
typed specialization
\citep{danvy-simple,Danvy-tagging-encoding,hughes-type,Birkedal-PE-ML,taha-tag,Makholm-TagElim}
and type systems (see~\S\ref{s:fancier}).

@@

This paper shows how to eliminate tagging overhead, whether in the
context of code generation and whether in the presence of dispatching
overhead.  We use metalanguage types, without such fancy features as
generalized algebraic data types (GADTs) or dependent types, to rule out
ill-typed object terms statically,
thus speeding up interpretation and assuring that our interpreters
do not get stuck.  We illustrate the problem of tagging overhead
in this section using a simple evaluator as example.  We apply our
solution first to evaluation, then to code\hyp generation tasks such as
partial evaluation.

@@

Conversely, the type system of the metalanguage checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checkers also flag syntactically invalid
object terms, such as if we forget |app| somewhere above.
Because our encoding of terms and types are so straightforward and
metacircular, these error messages from the metalanguage implementation
are just about as readable as those for ``native'' type errors such as
|fun x -> x x|.

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, we unfold |fix| rather than residualize it.
A conservative alternative is to unfold recursion only once, then residualize:
\begin{code}
let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
\end{code}
Many sophisticated approaches have been developed to decide how much to unfold
\citep{Jones-Mix,jones-partial}, but this issue is 
orthogonal to our presentation.
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{BondorfDanvy}.

@@

Our partial evaluator owes much to
\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducers for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen,Mogensen-SelfApplicable} also
build static and dynamic terms in tandem.  However, they
build a static term for every object term, even a bound variable,
so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remain untyped.  
In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.
The idea of generating static and dynamic components alongside each
other is part of the tradition that developed partial evaluators such as
Schism \citep[\S5]{consel-schism-93}.

@@

\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
apply HOAS to typed self\hyp interpretation.  However, their source
language is untyped.  Therefore, their object\hyp term encoding has
tags, and their interpreter can raise run-time errors.
Nevertheless, HOAS lets the partial
evaluator remove all the tags. In contrast, our object encoding and
interpreters do not have tags to start with and obviously cannot
raise run-time errors.
